Subgroup detection in linear growth curve models with GLMM trees

### 2023-06-22 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Version 1.2-2: Added experimental betamertree function
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3258 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6531527a0e69b00698bd34a78dca3b18408c10df
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3262 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 8b30d46ca33a3409789ca7731f27d1ba4c5f31fc

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

@@

The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and SEM trees based on likelihood ratio tests (LRTs). All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, some split criterion (such as a $p$-value from the corresponding LRT) is computed for every possible cutpoint. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection.

In summary, compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection.

In what follows, we first describe estimation of GLMM trees in general along with the novel adjustments geared towards longitudinal data structures. Subsequently, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees. In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and implications.

@@

\section{Estimation}
\label{sec:estimation}

@@

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML). Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm. The initialization simply assumes zero random effects.
%
\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and all random-effect estimates $\hat{b}_{i,(r)} = 0$.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset, yielding the partition $j_{(r)}$.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j, (r)} + Z_{i} b_{i, (r)}$ with the partition $j_{(r)}$ from Step~1. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}
%
Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j_{(r)}$ from the GLM tree is the same as $j_{(r-1)}$ from the previous step.

In the following two subsections provide some more details on alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respecitvely. Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.

@@

Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more closely than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy. More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
The alternative initialization step is thus:
%
\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\arabic{enumi}$'$.}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta + Z_{i} b_{i, (r)}$ to the full sample. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
\end{enumerate}

@@

\section{Simulation Study I}

We performed a simulation study to compare the performance GLM(M) trees estimated by initializing estimation with the subgroup structure versus the random effects, and employing observation- versus cluster-level covariances for the parameter stability tests.

@@

We applied ten variations LM(M) trees to every generated dataset, where each variation combines one of the following three \emph{random-effects specifications} (none/intercepts/intercepts+slopes) with the \emph{random-effect initializations} (if any; all zero vs.\ full sample estimates) and \emph{covariance specifications in the parameter instability tests} (classic vs.\ clustered), as discussed in the previous section.
%
\begin{itemize}
  \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications.
  \begin{itemize}
      \item Default: Classic observation-level covariances.
      \item Alternative: Clustered covariances. 
  \end{itemize}

  \item Intercepts: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$. This yields linear mixed-effects model (LMM) trees in which only the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0. The four variations considered are the following.
  \begin{itemize}
    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (0.). 
    \item Alternative: Clustered covariances in the parameter stability tests.
    \item Alternative: Random-effect initialization with the full-sample estimates (0$'$.). 
    \item Alternative: Clustered covariances and random-effect initialization with the full-sample estimates.
  \end{itemize}

  \item Intercepts and slopes: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$. This yields LMM trees in which the variance of both the random intercept and slope were freely estimated. The four variations considered are the same as for the intercept-only LMM trees.
\end{itemize}
%
To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}. To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}. To compute clustered covariance, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}. We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.

@@

We evaluated tree accuracy by counting the number of splits in each tree. Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables). We also assessed the variable selected for the first split in every tree.

@@

\emph{FIXME: Figure~2 and its caption are inconsistent because LM trees are not on the left but on the right. Also, the following paragraphs are inconsistent with respect to the discussion of MAD. The first occurrence of MAD is maybe M?}

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. Irrespective of the random-effect specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average. This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, irrespective of the random-effects specification. The average number of splits of LMM trees employing clustered covariances were closest to the true number of splits. In terms of mean absolute deviation (MAD) from the true tree size, LM trees with clustered covariances performed best, closely followed by the LMM trees employing clustered covariances.

In terms of MAD, however, the LMM tree estimating random intercepts only performed even slightly better than the latter two approaches.  Thus, initializing estimation with the random effects can successfully mitigate the overfitting observed for the default LMM trees, but the effect depends strongly on the random-effects specification. With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure. Use of random-effects initialization and cluster-level covariances in tandem was less effective in reducing the number of splits than using either of the two approaches.

Distributions of the number of splits, separated according to the levels of the data-generating design are presented and discussed in  Figure~\ref{fig:LMM_sizes_interact} and Appendix~\ref{sec:AppendixA}, because they show a pattern very similar to Figure~\ref{fig:LMM_sizes}. Use of cluster-level covariances provides the most accurate split recovery. Of the data-generating parameters, $\sigma_{b_1}$ and $N$ showed the strongest effects, with higher values resulting in a higher number of splits. The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, and were involved in minor interactions: With the default fitting approach, they had a positive effect on the number of implemented splits; with clustered covariance or random-effects initialization, they had a negative effect on the number of splits. The correlation between partitioning variables hardly affected the number of implemented splits.

@@

\emph{FIXME: The `Note` below Table~1 (and~2) is too close to the lower horizontal rule, making it hard to read. The order of of the rows should match the order of the panels in
Figure~2, possibly aided by horizontal rules (then each combination of sigmas only has to be spelled out once).}

@@

## Print the table
kable_styling(add_footnote(kable(splits[1:10, -1], format = "latex", booktabs = TRUE, 
                                 label = "first_splits", align = "llccccc", 
        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
        caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L),
  "\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\hat{\\sigma}_{b_0}$ and $\\hat{\\sigma}_{b_1}$ are the estimated standard deviations of the random intercept and slope, respectively.", 
  notation="none", threeparttable = TRUE, escape = FALSE),
  font_size = 11, full_width=FALSE)
@

@@

\section{Simulation Study II}

@@

\section{Application}

@@

Our analyses focus on predicting trajectories of reading, math and science abilities, based on baseline characteristics. Time was measured as the number of months since the baseline assessment. In order to obtain approximately linear trajectories, $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.

\emph{README: Are these power transformations for months by 1/2 or 2/3 standard in the literature? Should this be cited in some way or is this uncontroversial?}

@@

\subsubsection{Partitioning methods}

@@

\emph{FIXME: Maybe better clarify what `existing data' in the next sentence means. I think you are referring to data where we have to evaluate out-of-sample rather than with the `true' effects in-sample, right?}

To compare the performance of LM(M) trees on existing data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}. Neither SEM trees, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations. Writing custom functionality to allow for prediction was most straightforward for longRPart. Furthermore, performance of longRPart will be representative of the other methods, in the sense that all four methods fit full mixed-effects models locally. Like LongCART and longRPart2, longRPart fits a mixed-effects model in each of the terminal nodes. Like LRT-based SEM trees, longRPart uses an exhaustive LRT-based search in every node to find the optimal next split.

\emph{README: The conclusion not to consider SEM trees which performed rather well in the simulation studies is a bit unsatisfying. Also, longRPart was not included in the simulation studies. Overall, the new insights gained compared to the simulation studies are not so huge. So maybe we should discuss again how to best present this?}

Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student. In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default. This yielded a very large number of splits in the current analyses and poor predictive accuracy. We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits. Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but this has not been implemented.

@@

The ECLS-K datasets have exceptionally large sample sizes. We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology. For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions. We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition. Note that this measures the accuracy of the fixed-effects predictions only. Tree size was measured by the number of splits in each tree, and computation time was measured in seconds.

@@

\begin{figure}[!bt]
\caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=2, width=8>>=
## Load datasets for computing variance of response variables
load("Reading ability data.Rdata")
var_read <- var(readdata$score)
load("Math ability data.Rdata")
var_math <- var(mathdata$score)
load("Science ability data.Rdata")
var_scie <- var(sciedata$score)

## MSE
load("MSEs, N = 250, Math.Rda")
math_MSEs <- cbind(MSEs, N = 250)
load("MSEs, N = 1000, Math.Rda")
math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart
#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart_s
#sapply(math_MSEs, function(x) tapply(x, math_MSEs$N, mean, na.rm =TRUE))

>>>>

\title{Subgroup detection in linear growth curve models with GLMM trees}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

@@

The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, some split criterion (such as a $p$-value from a likelihood-ratio test) is computed for every possible cutpoint. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection.

Compared to other methods for partitioning LGCMs, GLMM trees may provide lower computational load, lower model complexity and instability, and unbiased variable selection. As such, GLMM trees could prove particularly useful for subgroup detection in longitudinal trajectories. In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments that may be particularly useful for partitioning longitudinal trajectories. We use a dataset on science ability trajectories te illustrate the relevance of the adjustments. Subsequently, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets. In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART, in simulated datasets. In the Empirical~Evaluation, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

@@

\section{Estimation of GLMM trees}
\label{sec:estimation}

@@

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML). Thus, when it is known into which node~$j$ each subject~$i$ falls, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm. The initialization simply assumes zero random effects.
%
\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and all random-effect estimates $\hat{b}_{i,(r)} = 0$.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset, yielding the partition $j_{(r)}$.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j, (r)} + Z_{i} b_{i, (r)}$ with the partition $j_{(r)}$ from Step~1. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}
%
Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j_{(r)}$ from the GLM tree is the same as $j_{(r-1)}$ from the previous step.

The following two subsections describe alternative initialization strategies for Step~0 and details of the GLM tree algorithm from Step~1, respectively. Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve the performance when partitioning LGCMs using longitudinal data.

@@

Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} specifically assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy. More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

The alternative initialization step is thus:
%
\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\arabic{enumi}$'$.}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta + Z_{i} b_{i, (r)}$ to the full sample. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
\end{enumerate}

@@

\subsection{Illustration on real data}

To illustrate the proposed adjustments on real data, we use a subsample of 250 children from a much larger dataset of the Early Childhood Longitudinal Study \citep[ECLS-K; ][]{NCES10}. Here, we partition trajectories of science ability, which were assessed in 3rd, 5th and 8th grade. We used 11 potential partitioning variables, including sociodemographics, motor and interpersonal skills. The dataset is described and analyzed in more detail in Study~III. With a sample of 250 children, we expect to obtain a relatively small number of relevant subgroups that can more easily be depicted and interpreted.

\begin{figure}
\caption{Trees resulting from various GLMM tree fitting approaches.}
\begin{subfigure}{1.1\textwidth}%
<<fig=TRUE, height=5.5, width=10, echo=FALSE>>=
library("glmertree")
load("sizes, N = 250, Science.Rda")
scie_sizes <- cbind(sizes, N = 250)
load("bag_ids, N = 250, Science.Rda")
load("Science ability data.Rdata")
data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
data$CHILDID <- factor(data$CHILDID)
data$GENDER <- factor(data$GENDER)
data$asmtmm <- factor(data$asmtmm)
LMM_form <- score ~ months | CHILDID | GENDER + RACE + WKSESL + C1GMOTOR + 
  C1FMOTOR + T1INTERN + T1EXTERN + T1INTERP + T1CONTRO + P1FIRKDG + AGEBASELINE

lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "tree", type = "simple", gp = gpar(cex = .7))
icc <- as.data.frame(VarCorr(lmm_tree))

testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from application of default settings.}
\end{subfigure}%

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
plot(lmm_tree_r, which = "tree", type = "simple", gp = gpar(cex = .7))
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of random-effects initialization.}
\end{subfigure}%

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
plot(lmm_tree_c, which = "tree", type = "simple", gp = gpar(cex = .7))
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of clustered covariances.}
\end{subfigure}%

\label{fig:LMM_trees}
\end{figure}

Using the the default fitting approach, the tree in the top panel of Figure~\ref{fig:LMM_trees} was obtained, comprising \Sexpr{(length(lmm_tree$tree)+1)/2} terminal nodes. Each node has its own fixed estimate for the intercept (science ability at baseline) and slope of time in months.  The most important partitioning variable is a measure for socio-economic backgroud (WKSESL), with lower values yielding lower values for the intercept, indicating that children with lower SES score tend to have lower science skills both at the baseline and later assessments. Further, we see effects of internalizing problems (T1INTERN), gender, race and fine motor skills (C1FMOTOR), which we do not inspect in detail here as we want to focus on cahnges in results with the proposed adjustments to the estimation approach. The proportion of variance in science skill scores explained in science ability scores by the tree is \Sexpr{round(1 - mean((lmm_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, when evaluated on observations not part of the training sample. %The intra-class correlation for the random intercept term is \Sexpr{icc$vcov[1]/sum(icc$vcov)}.

The middle panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, by initializing estimation with the random effects. A slightly different set of partitioning variables were selected and the tree has less terminal nodes. The percentage of variance in science skill scores explained by the tree is \Sexpr{round(1 - mean((lmm_r_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, slightly lower than for the GLMM tree fitted with default settings. %The intra-class correlation for the random intercept term is \Sexpr{round(icc_r$vcov[1]/sum(icc_r$vcov), digits = 2)}, slightly higher than for the previous GLMM tree. 
The bottom panel of Figure~\ref{fig:LMM_trees} depicts the GLMM tree fitted on the science trajectories, estimated using clustered covariances. The resulting tree is the smallest tree obtained, and selected SES and internalizing problems for partitioning. With a proportion of variance explained of \Sexpr{round(1 - mean((lmm_c_preds - testdat$score)^2) / var(testdat$score), digits = 3)}, providing highest predictive accuracy of the three fitting approaches. %The intra-class correlation for the random intercept term is \Sexpr{round(icc_c$vcov[1]/sum(icc_c$vcov), digits = 3)}, higher than obtained with the previous two GLMM trees.

\section{Study~1}

@@

We applied ten variations LM(M) trees to every generated dataset, where each variation combines one of the following three \emph{random-effects specifications} (none/intercepts/intercepts+slopes) with the \emph{random-effect initializations} (if any; all zero vs.\ full sample estimates) and \emph{covariance specifications in the parameter instability tests} (classic vs.\ clustered), as discussed in the previous section.
%
\begin{itemize}
  \item None: $\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$. Random effects and thus their variance were fixed to 0, yielding linear model (LM) trees with fixed effects only and the following variations of covariance specifications:
  \begin{itemize}
      \item Default: Classic observation-level covariances.
      \item Alternative: Clustered covariances. 
  \end{itemize}

  \item Intercepts: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} = 0$. This yields linear mixed-effects model (LMM) trees in which only the variance of the random intercept was freely estimated and the variance of the random slope was fixed to 0. The four variations considered are the following:
  \begin{itemize}
    \item Default: Classic observation-level covariances in the parameter stability tests and random-effect initialization with all zeros (0.). 
    \item Alternative: Clustered covariances in the parameter stability tests.
    \item Alternative: Random-effect initialization with the full-sample estimates (0$'$.). 
    \item Alternative: Clustered covariances and random-effect initialization with the full-sample estimates.
  \end{itemize}

  \item Intercepts and slopes: $\hat{\sigma}_{b_0} > 0; \hat{\sigma}_{b_1} > 0$. This yields LMM trees in which the variance of both the random intercept and slope were freely estimated. The four variations considered are the same as for the intercept-only LMM trees.
\end{itemize}
%
To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}. To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}. To compute clustered covariances, we used package \textbf{sandwich} \citep[version 3.0-1; ][]{ZeilyKoll20}. We employed outer product of gradients method to compute covariances, thus employing only the meat of the sandwich estimator to estimate the covariances.

@@

We evaluated tree accuracy by counting the number of splits in each tree, and computing mean absolute deviation (MAD) from the true tree size (3 splits). Trees with $> 3$ splits are indicative of Type-I error, while trees with $< 3$ splits are indicative of Type-II errors (i.e., too low power for tests of the true partitioning variables). We also assessed the variable selected for the first split in every tree.

@@

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. Irrespective of random-effects specification, the default fitting approaches (left-most box in every panel) tended to overfit, implementing $> 3$ splits in \Sexpr{format(100*mean(c(props_more_splits["lm"], props_more_splits["lmer"], props_more_splits["lmer_s"])), digits = 2L)}\% datasets, on average. This overfitting was successfully mitigated through use of cluster-level covariances in the parameter stability tests, again irrespective of the random-effects specification. The average number of splits in LMM trees employing clustered covariances was closest to the true number of splits. In terms of MAD, however, the LMM tree initializing estimation with the random effects, when only a random intercept was freely estimated, performed best. Thus, initializing estimation with the random effects can successfully mitigate overfitting observed for the default LMM trees, but this effect depends strongly on the random-effects specification. With estimation of both a random intercept and slope, random-effects initialization may lead to subgroup differences being captured by the random effects instead of the tree structure.

Irrespective of the random-effects specification, combining the use random-effects initialization and cluster-level covariances was not very effective. Use of clustered covariances seems to provide the most reliably improvement in performance.

Distributions of the number of splits, separated according to the levels of the data-generating design are omitted here because this revealed a pattern very similar to Figure~\ref{fig:LMM_sizes}. Thus, the optimal fitting approach did not seem dependent on data characteristics.

These results are presented Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}. Of the data-generating parameters, $\sigma_{b_1}$ and $N$ showed the strongest effects, with higher values resulting in a higher number of splits. The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, and were involved in minor interactions: With the default fitting approach, they had a positive effect on the number of implemented splits; with clustered covariance or random-effects initialization, they had a negative effect on the number of splits. The correlation between partitioning variables hardly affected the number of implemented splits.

@@

@@

## Print the table
kable_styling(add_footnote(row_spec(kable(splits[1:10, -1], format = "latex", booktabs = TRUE, 
                                 label = "first_splits", align = "llccccc", 
        escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
        caption = "Variables selected for the first split by each LM tree (top two rows) and LMM tree (bottom eight rows) estimation approach.", row.names = FALSE, digits = 3L), row = c(2, 6), hline_after =  TRUE),
  "\\textit{Note.} $U_1$ is the true first splitting variable and is binary; all other partitioning variables are continuous, with $U_2$ and $U_3$ being true splitting variables (nodes 2 and 3). $\\hat{\\sigma}_{b_0}$ and $\\hat{\\sigma}_{b_1}$ are the estimated standard deviations of the random intercept and slope, respectively.", 
  notation="none", threeparttable = TRUE, escape = FALSE),
  font_size = 11, full_width=FALSE)
@

@@

\section{Study~II}

@@

\section{Study~III}

@@

Our analyses focus on predicting trajectories of reading, math and science abilities, based on baseline characteristics. Time was measured as the number of months since the baseline assessment. In order to obtain approximately linear trajectories, we chose metrics of time based on visual inspection of the observed data: $\mathrm{months}^{1/2}$ was used as the timing metric for reading and math trajectories, and $\mathrm{months}^{2/3}$ for science trajectories.

@@

\subsubsection{Fitting approaches}

@@

%To compare performance of LM(M) trees on real data, we could not employ SEM tree or LongCART, so we opted for longRPart \citep{AbdoyLeBl02}. Neither SEM trees, LongCART, longRPart or longRPart2 implementations provide functionality to obtain predictions for new observations. Writing custom functionality to allow for prediction was most straightforward for longRPart. Furthermore, performance of longRPart will be representative of the other methods, in the sense that all four methods fit full mixed-effects models locally. Like LongCART and longRPart2, longRPart fits a mixed-effects model in each of the terminal nodes. Like LRT-based SEM trees, longRPart uses an exhaustive LRT-based search in every node to find the optimal next split.

%Using package \textbf{longRPart} \citep[version 1.0; ][]{StewyAbdo12}, we fitted two trees with different random effects specifications: A tree comprising a random intercept only, and a tree comprising both a random intercept and slope with respect to student. In order to implement a split, longRPart uses a criterion based on the minimum decrease in deviance, which is set to 0.01, by default. This yielded a very large number of splits in the current analyses and poor predictive accuracy. We therefore restricted maximum tree depth to 5, yielding a maximum of $2^5$ = \Sexpr{2^5} terminal nodes, or \Sexpr{(2^5)-1} splits. Significance of the LRT, or a cross-validation approach to determine whether to implement a split or not may yield better results, but this has not been implemented.

@@

The ECLS-K datasets have exceptionally large sample sizes. We therefore employed random sampling to obtain training samples of $N=250$ children, likely more representative of real-world studies in psychology. For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions. We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition. Note that this measures the accuracy of the fixed-effects predictions only. Tree size was measured by the number of splits in each tree.

@@

\begin{figure}[!bt]
\caption{Mean squared errors for trees fitted to math (top), reading (middle) and science (bottom) ability trajectories.}
\begin{subfigure}{\textwidth}
<<echo=FALSE, fig=TRUE, height=2, width=8>>=
## Load datasets for computing variance of response variables
load("Reading ability data.Rdata")
var_read <- var(readdata$score)
load("Math ability data.Rdata")
var_math <- var(mathdata$score)
load("Science ability data.Rdata")
var_scie <- var(sciedata$score)

## MSE
load("MSEs, N = 250, Math.Rda")
math_MSEs <- cbind(MSEs, N = 250)
#load("MSEs, N = 1000, Math.Rda")
#math_MSEs <- rbind(math_MSEs, cbind(MSEs, N = 1000))
#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart
#sapply(math_MSEs, function(x) table(is.na(x)))$longRPart_s
#sapply(math_MSEs, function(x) tapply(x, math_MSEs$N, mean, na.rm =TRUE))
